Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-04-29 paper/ms.tex
update ms

%%%%%%%%%%%%%
% OFT-CITED %
%%%%%%%%%%%%%
\defcitealias{McQuillan_2014}{M14}
\defcitealias{Mazeh_2015}{M15}
\defcitealias{Santos_2019}{S19}
\defcitealias{Santos_2021}{S21}
\defcitealias{Fulton_2018}{F18}
\defcitealias{Berger_2020a_catalog}{B20}

@@

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  Today
however, measurements of stellar rotation periods in open clusters
provide the most accurate empirical calibrations of rotation-based
ages (\citealt{Bouma_2023}, and references therein).  This approach
ultimately ties the rotation age scale to the cluster age scale, and
is limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.  The oldest
cluster for which this method has been calibrated is currently M67
\citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.

Lithium ages includes two qualitatively distinct regimes: depletion
boundary ages derived for M-dwarfs in star clusters, and the less
precise decline-based ages for individual field FGK stars
\citep{Soderblom_2010}.  Here we focus on the latter approach, which
relies on the observation that the Li abundances of
partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
Empirical understanding has however recently improved due to the work
by \citet{Jeffries_2023}, who built a model for how the equivalent
width (EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function
of stellar effective temperature and age for a set of 6{,}200 stars in
52 open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

>>>>

%%%%%%%%%%%%%
% OFT-CITED %
%%%%%%%%%%%%%
\defcitealias{2013ApJ...775L..11M}{M13}
\defcitealias{McQuillan_2014}{M14}
\defcitealias{Mazeh_2015}{M15}
\defcitealias{Santos_2019}{S19}
\defcitealias{Santos_2021}{S21}
\defcitealias{Fulton_2018}{F18}
\defcitealias{Berger_2020a_catalog}{B20}

@@

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-4\,Gyr, and can also provide useful age limits at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
In the context of Kepler,
empirical models have been previously applied to derive rotation-based
ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015}.
There have been major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)}]
  \item The depth of the Kepler Object of Interest catalog, as well as our understanding of which of these objects are
    false positives;
  %
  \item Our empirical understanding of how rotation periods of stars
    in open clusters evolve
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation periods as a function of time
    \citep{Bouma_2023}.  
    The new model is broadly consistent with older calibrations
    \citep[e.g.][]{Mamajek_2008} for Sun-like stars.  However, for
    late G dwarfs and K dwarfs, this age-calibration differs by
    factors of a few from previous estimates;
  %
  \item Our ability to identify binary stars, which can potentially
    problematic for estimates of rotation-based ages.
\end{enumerate}
Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
empirically tied to the cluster age scale and rotation sequence, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

Beyond improvements in rotation-based ages, recent years have also
yielded improvements in the lithium age scale.
Lithium ages include two qualitatively distinct regimes: depletion
boundary ages derived for M-dwarfs in star clusters, and the less
precise decline-based ages for individual field FGK stars
\citep{Soderblom_2010}.  Here we focus on the latter approach, which
relies on the observation that the Li abundances of
partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
Empirical understanding has however recently improved due to the work
by \citet{Jeffries_2023}, who built a model for how the equivalent
width (EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function
of stellar effective temperature and age for a set of 6{,}200 stars in
52 open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.
